---
title: "GHG Inventory: Emissions from Manure Management"
author: "*Amy Plambeck*"
date: "8/25/2020"
output: 
  html_document:
    toc: TRUE
    toc_float: TRUE
    df_print: paged
    code_download: true
    theme: "cosmo"
---



```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, error = TRUE, message = FALSE, warning = FALSE)
```

```{r, results= "hide",warning=FALSE, message=FALSE}
library(tidyverse)
library(sf)
library(lsr)
library(readxl)
library(dplyr, warn.conflicts = FALSE)

# Suppress summarise info
options(dplyr.summarise.inform = FALSE)
```

# 1. Data, Part 1 (Sorting)

## 1.1 Importing source data

MPCA Manure Management Data

```{r}
mpca_source <- read_excel("C:/Users/PlambeA/Metropolitan Council/Leon, Mauricio - 01_ghg_inventory/city_township/01_data/0106_livestock_and_agriculture/010601_livestock/01060102_county/mpca_manure_datareq_jun2020/Feedlot registrations_MPCA_20200617.xlsx", 
    sheet = "Feedlot Warehouse_filtered")


```

WMS and Methane Emissions Conversion Factors
```{r}
wms_conv <- read_excel("livestock_inv_manure_param.xlsx", 
    sheet = "wms_conversionfactor")
```

Animals Characteristics (VS rates, methane emissions potential, typical animal mass)
```{r}
animal_char <- read_excel("livestock_inv_manure_param.xlsx", 
    sheet = "animal_characteristics")
```

## 1.2 Create flags for all cattle and feedlot/NOF cattle

Cattle data uses different methodology. Can be both liquid or solid. 

Mutate cattle indicator (currently includes `Beef Calf`|`Bull Mature` but those with pasture flag = 0 will omits NOF Calves and NOF Bulls per animal_char)
```{r}
mpca_source<-
  mpca_source %>% 
  mutate(cattle_flag = ifelse(`Beef Heifer` | `Beef Steer Stock`| `Dairy Cattle` | `Dairy Cattle Little` | `Dairy Heifer`| `Beef Cow Calf`|`Beef Calf`|`Bull Mature`|`Veal Calf` > 0, 1, 0))
```

```{r}
mpca_source<-
  mpca_source %>% 
  mutate("feedlot_cattle_flag" = ifelse(cattle_flag ==1 & `Pasture Flag` == "N", 0, 1))
```

Note here that this is a step in how we produce compatibility labels in the animal_char spreadsheet. Step 1.5 and 1.6 also deal with the compatibility labels (compat_lab).


## 1.3 Create flags for swine
Swine manure is always liquid, so we'll want to make sure it's not counted as solid when there are both liquid and solid storage on site. 
```{r}
mpca_source<-
  mpca_source %>% 
  mutate(swine_flag = ifelse(`Swine Big`|`Swine Little`|`Swine Medium` > 0, 1, 0))
```

## 1.4 Create flags for poultry
```{r}
mpca_source<-
  mpca_source %>% 
  mutate(poultry_flag = ifelse(`Turkey Big`|`Turkey Little`| `Peacocks`|`Geese`|`Fowl`|`Chicken Broiler Big`|`Chicken Broiler Little`|`Chicken Layer Big`|`Chicken Layer Little` > 0, 1, 0))
```

## 1.5 Combine "little" and "big" totals

```{r}
mpca_source<-
  mpca_source %>% 
  mutate(
    "Swine All" = `Swine Big` + `Swine Little` + `Swine Medium`, na.rm = TRUE,
    "Turkey All" = `Turkey Big`+`Turkey Little`,na.rm = TRUE,
    "Broiler All" = `Chicken Broiler Big` + `Chicken Broiler Little`, na.rm = TRUE,
    "Layer All" = `Chicken Layer Big` + `Chicken Layer Little`, na.rm = TRUE,
    "Dairy Cows All" = `Dairy Cattle` + `Dairy Cattle Little`, na.rm = TRUE
    )
```  

## 1.6 Designate fields for feedlot and NOF groups of cattle

```{r}
mpca_source<-
  mpca_source %>% 
  mutate(
    "Feedlot Steer"= ifelse(feedlot_cattle_flag == 1, `Beef Steer Stock`, NA), 
    "Feedlot Heifers" = ifelse(feedlot_cattle_flag == 1, `Beef Heifer`, NA),
    "NOF Bulls" = ifelse(feedlot_cattle_flag == 0, `Bull Mature`, NA), 
    "NOF Calves" = ifelse(feedlot_cattle_flag == 0, `Beef Calf`, NA),
    "NOF Heifers" = ifelse(feedlot_cattle_flag == 0, `Beef Heifer`, NA),
    "NOF Steers" = ifelse(feedlot_cattle_flag == 0, `Beef Steer Stock`, NA), 
    "NOF Cows" = ifelse(feedlot_cattle_flag == 0, `Beef Cow Calf`, NA)
  )
```


## 1.7 Sort data into poultry and liquid, solid, and both


Poultry - Poultry are not subject to liquid/solid waste management conversion factors. 

```{r}
poultry <-
  mpca_source %>% 
  filter(poultry_flag == 1)
```


Liquid Storage - Explicitly liquid waste management on site only
```{r}
liquid_storage <-
  mpca_source %>% 
  filter(`Liquid Storage Flag` == "Y" | swine_flag ==1 | poultry_flag==0) %>% 
  filter(`Solid Storage Flag` == "N"|NA) 
```

Solid Storage - Explicitly solid waste management on site only
```{r}
solid_storage <-
  mpca_source %>% 
  filter(`Solid Storage Flag` == "Y" | swine_flag == 0) %>% 
  filter(`Liquid Storage Flag` == "N"|NA)
```

Liquid and Solid Storage - Both solid and liquid waste management on site, distinguish by deduction?
```{r}
liquid_and_solid_storage <-
  mpca_source %>% 
  filter(`Solid Storage Flag` == "Y" & `Liquid Storage Flag` == "Y")
  
```

Pasture only - because the WMS conversion is so low and the limitations through this data set to sort how much of the or what kind of waste produced is going to pasture, this will be the data set for when Pasture = Y and Liquid & Solid = N
```{r}
pasture_storage <-
  mpca_source %>% 
  filter(`Solid Storage Flag` == "N" & `Liquid Storage Flag` == "N" & `Pasture Flag` == "Y")
```


# 2. Data, Part 2 (Joining)

## 2.1 Bring together Animal Counts by County

```{r}
county_poultry <- poultry %>% 
  group_by(`Site County Name`) %>% 
  summarize(
    `Broiler All` = sum(na.exclude(`Broiler All`)),
    `Layer All` = sum(na.exclude(`Layer All`)),
    `Turkey All` = sum(na.exclude(`Turkey All`))
  )
```


```{r}
county_solid<- solid_storage %>% 
  group_by(`Site County Name`) %>% 
  summarise(
    `Feedlot Steer`=sum(na.exclude(`Feedlot Steer`)), 
    `Dairy Cows All` = sum(na.exclude(`Dairy Cows All`)),
    `Dairy Heifer` = sum(na.exclude(`Dairy Heifer`)),
    `Feedlot Heifers` = sum(na.exclude(`Feedlot Heifers`)),
    `NOF Bulls` = sum(na.exclude(`NOF Bulls`)),
    `NOF Calves` = sum(na.exclude(`NOF Calves`)),
    `NOF Heifers` = sum(na.exclude(`NOF Heifers`)),
    `NOF Steers` = sum(na.exclude(`NOF Steers`)),
    `NOF Cows` = sum(na.exclude(`NOF Cows`)),
    `Swine All`  = sum(na.exclude(`Swine All`)),
    `Sheep` = sum(na.exclude(`Sheep`)),
    `Goats` = sum(na.exclude(`Goats`)),
    `Horse`  = sum(na.exclude(`Horse`))
    
    )

```

```{r}
county_liquid<- liquid_storage %>% 
  group_by(`Site County Name`) %>% 
  summarise(
    `Feedlot Steer`=sum(na.exclude(`Feedlot Steer`)), 
    `Dairy Cows All` = sum(na.exclude(`Dairy Cows All`)),
    `Dairy Heifer` = sum(na.exclude(`Dairy Heifer`)),
    `Feedlot Heifers` = sum(na.exclude(`Feedlot Heifers`)),
    `NOF Bulls` = sum(na.exclude(`NOF Bulls`)),
    `NOF Calves` = sum(na.exclude(`NOF Calves`)),
    `NOF Heifers` = sum(na.exclude(`NOF Heifers`)),
    `NOF Steers` = sum(na.exclude(`NOF Steers`)),
    `NOF Cows` = sum(na.exclude(`NOF Cows`)),
    `Swine All`  = sum(na.exclude(`Swine All`)),
    `Sheep` = sum(na.exclude(`Sheep`)),
    `Goats` = sum(na.exclude(`Goats`)),
    `Horse`  = sum(na.exclude(`Horse`))
    
    )
```

```{r}
county_liquid_and_solid<- liquid_and_solid_storage %>% 
  group_by(`Site County Name`) %>% 
  summarise(
    `Feedlot Steer`=sum(na.exclude(`Feedlot Steer`)), 
    `Dairy Cows All` = sum(na.exclude(`Dairy Cows All`)),
    `Dairy Heifer` = sum(na.exclude(`Dairy Heifer`)),
    `Feedlot Heifers` = sum(na.exclude(`Feedlot Heifers`)),
    `NOF Bulls` = sum(na.exclude(`NOF Bulls`)),
    `NOF Calves` = sum(na.exclude(`NOF Calves`)),
    `NOF Heifers` = sum(na.exclude(`NOF Heifers`)),
    `NOF Steers` = sum(na.exclude(`NOF Steers`)),
    `NOF Cows` = sum(na.exclude(`NOF Cows`)),
    `Swine All`  = sum(na.exclude(`Swine All`)),
    `Sheep` = sum(na.exclude(`Sheep`)),
    `Goats` = sum(na.exclude(`Goats`)),
    `Horse`  = sum(na.exclude(`Horse`))
    
    )
```

```{r}
county_pasture<- pasture_storage %>% 
  group_by(`Site County Name`) %>% 
  summarise(
    `Feedlot Steer`=sum(na.exclude(`Feedlot Steer`)), 
    `Dairy Cows All` = sum(na.exclude(`Dairy Cows All`)),
    `Dairy Heifer` = sum(na.exclude(`Dairy Heifer`)),
    `Feedlot Heifers` = sum(na.exclude(`Feedlot Heifers`)),
    `NOF Bulls` = sum(na.exclude(`NOF Bulls`)),
    `NOF Calves` = sum(na.exclude(`NOF Calves`)),
    `NOF Heifers` = sum(na.exclude(`NOF Heifers`)),
    `NOF Steers` = sum(na.exclude(`NOF Steers`)),
    `NOF Cows` = sum(na.exclude(`NOF Cows`)),
    `Swine All`  = sum(na.exclude(`Swine All`)),
    `Sheep` = sum(na.exclude(`Sheep`)),
    `Goats` = sum(na.exclude(`Goats`)),
    `Horse`  = sum(na.exclude(`Horse`))
    
    )
```


## 2.2 Merge animal conversion factors to County

Gather counties into into long data set

```{r}
county_poultry <- county_poultry %>% 
  gather(key = "animal", value = "head count", 2:4)

county_solid <- county_solid %>%
  gather(key = "animal", value = "head count", 2:14) 

county_liquid <- county_liquid %>%
  gather(key = "animal", value = "head count", 2:14) 

county_liquid_and_solid <- county_liquid_and_solid %>%
  gather(key = "animal", value = "head count", 2:14) 

county_pasture <- county_pasture %>% 
  gather (key = "animal", value = "head count", 2:14)
```

Join the first set of conversion factors to the head count by county data sets from above
```{r}
co_poultry <- right_join(county_poultry, animal_char, by = c("animal" = "compat_lab"))

co_solid <- right_join(county_solid, animal_char, by = c("animal" = "compat_lab"))

co_liquid <- right_join(county_liquid, animal_char, by = c("animal" = "compat_lab"))

co_liquidsolid <- right_join(county_liquid_and_solid, animal_char, by = c("animal" = "compat_lab"))

co_pasture <- right_join(county_pasture, animal_char, by = c("animal" = "compat_lab"))
```

## 2.3 Sort out cattle, "non-cattle" so to speak

Most non-cattle livestock have VS production measured in VS per day per animal unit (being 1,000 lbs), with the exception of NOF Bulls and Calves who are also measured in this way. Most cattle have their VS production measured in VS per year per head-- no weight involved. 

vs/day/au = all animals + nof bull, nof calves
vs/year/head = all cattle except nof bulls, nof calves

Create tags for vs/day and vs/year in the form of variable "VS_measure".

```{r}
co_poultry <- 
  co_poultry %>% 
  mutate(VS_measure = "vsDay" ) %>% 
  filter(`head count` != 0|NA)
  
co_solid <- 
  co_solid %>% 
  mutate(VS_measure = ifelse(animal == 'Swine All'|animal =='Goats'|animal =='Horse'|animal =='NOF Bulls'|animal =='NOF Calves', "vsDay", "vsYear" )) %>% 
  filter(`head count` != 0|NA)
         
co_liquid<-
  co_liquid %>% 
  mutate(VS_measure = ifelse(animal == 'Swine All'| animal =='Sheep'|animal =='Goats'|animal =='Horse'|animal =='NOF Bulls'|animal =='NOF Calves', "vsDay", "vsYear" )) %>% 
  filter(`head count` != 0|NA)
         
co_liquidsolid <-
  co_liquidsolid %>% 
  mutate(VS_measure = ifelse(animal == 'Swine All'|animal =='Sheep'|animal =='Goats'|animal =='Horse'|animal =='NOF Bulls'|animal =='NOF Calves', "vsDay", "vsYear" )) %>% 
  filter(`head count` != 0|NA)
```


```{r}
co_pasture <-
  co_pasture %>% 
  mutate(VS_measure = ifelse(animal == 'Swine All' |animal == 'Sheep'|animal =='Goats'|animal =='Horse'|animal =='NOF Bulls'|animal =='NOF Calves', "vsDay", "vsYear" )) %>% 
  filter(`head count` != 0|NA)
```


## 2.4 Revising co_liquidsolid 

There are some certainties in this ambiguous set. Swine manure is always managed as liquid per the contact at MPCA, transfer to the appropriate place in co_liquid:

```{r}
swine <- filter(co_liquidsolid, animal_type== "Swine")

co_liquid<-co_liquid %>% 
  bind_rows(swine) %>% 
  group_by(`Site County Name`, animal, animal_type, tam, vs_rate_year_animal, vs_rate_day_au, bo, VS_measure) %>% 
  summarize(`head count` = sum(`head count`))

  
```

```{r}
co_liquidsolid <- filter(co_liquidsolid, animal_type!="Swine")
```



# 3. Calculations

Solid and liquid are relatively straightforward. Poultry, solid + liquid, and pasture sites are a bit more complicated for different reasons. 

Echoing above explanation: 

*Most non-cattle livestock have VS production measured in VS per day per animal unit (being 1,000 lbs), with the exception of NOF Bulls and Calves who are also measured in this way. Most cattle have their VS production measured in VS per year per head-- no weight involved. * 

Poultry also have different methane conversion factors, which will complicate how calculations at that step proceed.

The formula for calculating non-cattle livestock VS (with the two exceptions) will be

VS = Animal Population x (WMS*) x TAM /1000 x VS Rate x 365.25 

...and the VS formula for cattle will be

 VS = Animal Population x VS Rate x (WMS*)
 
*Note: In the ICLEI protocol the WMS variable is defined as the percentage of waste put toward a specific waste management system (thus WMS). Since this analysis splits up the different WMS into different datasets, the value is assumed to be 100% and so it is not necessary.

The formula for estimating methane emissions will be 

CH4 Emissions = VS * Bo * MCF * 0.662 x (1/1000) x GWP

GWP is 21, and the MCF is the methane conversion factor per WMS stored later in step 3.2.


 
## 3.1 Calculating volatile solid production from animals (kilograms per year)

```{r}

co_poultry <- co_poultry %>% 
  mutate(
    vsday = `head count` * `tam`/1000 * `vs_rate_day_au` * 365.25,
    vsyear = `head count` * `vs_rate_year_animal` 
    ) %>% 
  drop_na(`Site County Name`)

co_solid <- co_solid %>% 
  mutate(vsday = `head count` * `tam`/1000 * `vs_rate_day_au` * 365.25,
         vsyear = `head count` * `vs_rate_year_animal` 
         ) %>% 
  drop_na(`Site County Name`)

co_liquid <- co_liquid %>% 
  mutate(vsday = `head count` * `tam`/1000 * `vs_rate_day_au` * 365.25,
         vsyear = `head count` * `vs_rate_year_animal` 
         ) %>% 
  drop_na(`Site County Name`)
  
co_pasture <- co_pasture %>% 
  mutate(vsday = `head count` * `tam`/1000 * `vs_rate_day_au` * 365.25,
         vsyear = `head count` * `vs_rate_year_animal` 
         ) %>% 
  drop_na(`Site County Name`)

co_liquidsolid <- co_liquidsolid %>% 
  mutate(vsday = `head count` * `tam`/1000 * `vs_rate_day_au` * 365.25,
         vsyear = `head count` * `vs_rate_year_animal` 
         ) %>% 
  drop_na(`Site County Name`)

```

Combine the columns for vsday and vsyear (both represent VS produced in a year, just between the animals where conversion factors are in years or days. Slightly confusing! But they're comparable values.)

```{r}
co_pasture <- co_pasture %>% 
  mutate(vstotal = ifelse(is.na(vsday)==FALSE, vsday,
                          ifelse(is.na(vsyear)==FALSE, vsyear, NA)))

co_solid <- co_pasture %>% 
  mutate(vstotal = ifelse(is.na(vsday)==FALSE, vsday,
                          ifelse(is.na(vsyear)==FALSE, vsyear, NA)))

co_poultry <- co_pasture %>% 
  mutate(vstotal = ifelse(is.na(vsday)==FALSE, vsday,
                          ifelse(is.na(vsyear)==FALSE, vsyear, NA)))
co_liquid <- co_liquid %>% 
  mutate(vstotal = ifelse(is.na(vsday)==FALSE, vsday,
                          ifelse(is.na(vsyear)==FALSE, vsyear, NA)))


co_liquidsolid<- co_liquidsolid %>% 
  mutate(vstotal = ifelse(is.na(vsday)==FALSE, vsday,
                          ifelse(is.na(vsyear)==FALSE, vsyear, NA)))

```

## 3.2 Store WMS specific methane conversion factors as values for ease of use

```{r}
pasture <- (wms_conv$`conv_factor_perc`[1])
liquid <- (wms_conv$`conv_factor_perc`[2])
solid <- (wms_conv$`conv_factor_perc`[3])
poultry <- (wms_conv$`conv_factor_perc`[4])
```

## 3.3 Estimating all methane emissions in CO2e by county (*single* storage, excluding ambiguous storage) 

CH4 Emissions = VS * Bo * MCF * 0.662 x (1/1000) x GWP

GWP is 21, and the MCF is the methane conversion factor per WMS stored later in step 3.2.

**All of these will be summed by county!** 


```{r}
# Only run this chunk to keep WMS and animals separate

co_solid <- co_solid %>%
  mutate(ch4 = vstotal * bo * solid * 0.662 * (1/1000) * 21)  

co_liquid <- co_liquid %>%
  mutate(ch4 = vstotal * bo * liquid * 0.662 * (1/1000) * 21)

co_poultry <- co_poultry %>%
  mutate(ch4 = vstotal * bo * poultry * 0.662 * (1/1000) * 21)

co_pasture <- co_pasture %>%
  mutate(ch4 = vstotal * bo * pasture * 0.662 * (1/1000) * 21)

```

```{r}
county_ch4_unamb <- co_solid %>% 
  bind_rows(co_liquid, co_poultry,co_pasture) %>% 
  group_by(`Site County Name`) %>% 
  summarize(vstotal = sum(vstotal), ch4 = sum(ch4))

```


## 3.3 (1/2)
Just county level data is a bit restricted, so code and comments on how to 

  - Keep WMS separate 
  - Keep animal types separate
  - Both separate

have been kept. 

Just put the {r} back in to run.

Additional code: 

```{}
# Run the first chunk keeping WMS and animals separate and then this one to keep animals separate but combine WMS

county_ch4_unambig_animal_separate <- co_solid %>% 
  bind_rows(co_liquid, co_poultry,co_pasture) %>% 
  group_by(`Site County Name`,animal_type, animal) %>% 
  summarize(vstotal = sum(vstotal), ch4 = sum(ch4))

```

```{}
# Only run this chunk to keep WMS separate but not animals
co_solid <- co_solid %>%
  mutate(ch4 = vstotal * bo * solid * 0.662 * (1/1000) * 21) %>% 
  group_by(`Site County Name`) %>% 
  summarize(vstotal = sum(vstotal), ch4 = sum(ch4)) 
  

co_liquid <- co_liquid %>%
  mutate(ch4 = vstotal * bo * liquid * 0.662 * (1/1000) * 21) %>% 
  group_by(`Site County Name`) %>% 
  summarize(vstotal = sum(vstotal), ch4 = sum(ch4))

co_poultry <- co_poultry %>%
  mutate(ch4 = vstotal * bo * poultry * 0.662 * (1/1000) * 21) %>% 
  group_by(`Site County Name`) %>% 
  summarize(vstotal = sum(vstotal), ch4 = sum(ch4))

co_pasture <- co_pasture %>%
  mutate(ch4 = vstotal * bo * pasture * 0.662 * (1/1000) * 21) %>% 
  group_by(`Site County Name`) %>% 
  summarize(vstotal = sum(vstotal), ch4 = sum(ch4))
```


## 3.4 Estimating methane emissions in CO2e by county (*ambiguous* storage) 

**FINAL STEP**

Since there's no reliable way to know the manure management system details in this table, instead of a direct estimate of CH4 emissions we will create a range. Methane conversion factor values are 1 for pasture, 2 for solid, and 25 for liquid. This means there is a significant range of possible values.

Included is a mean CH4 estimate, calculated using the mean of the MCFs. This does not reflect any literature nor any real world data of livestock to WMS ratios. 

```{r}
co_liquidsolid <- co_liquidsolid %>%
  mutate(ch4_pasture = vstotal * bo * pasture * 0.662 * (1/1000) * 21, 
            ch4_solid = vstotal * bo * solid * 0.662 * (1/1000) * 21, 
                ch4_liquid = vstotal * bo * liquid * 0.662 * (1/1000) * 21, 
                   ch4_mean = vstotal * bo * 9 * 0.662 * (1/1000) * 21 ) %>% 
  group_by(`Site County Name`) %>% 
  summarize(vstotal = sum(vstotal), ch4_pasture = sum(ch4_pasture), ch4_solid = sum(ch4_solid), ch4_liquid = sum(ch4_liquid), ch4_mean=sum(ch4_mean))

```


```{r}
county_ch4 <- co_solid %>% 
  bind_rows(co_liquid, co_poultry,co_pasture,co_liquidsolid) %>% 
  group_by(`Site County Name`) %>% 
  summarize(vstotal = sum(vstotal), ch4 = sum(na.exclude(ch4)), addtl_ch4_if_pasture = sum(na.exclude(ch4_pasture)), addtl_ch4_if_solid = sum(na.exclude(ch4_solid)), addtl_ch4_if_liquid = sum(na.exclude(ch4_liquid)), addtl_ch4_if_mean = sum(na.exclude(ch4_mean))) %>% 
  group_by(`Site County Name`) %>% 
  mutate(total_ch4_low_est = sum(ch4, addtl_ch4_if_pasture),
         total_ch4_mean_est = sum(ch4, addtl_ch4_if_mean),
         total_ch4_high_est = sum(ch4, addtl_ch4_if_liquid))
```

## 3.5 View Final

```{r}
county_ch4
```



# Notes and Discussion

How can we improve the high/low estimate? Both poles -- all liquid and all pasture manure management-- are very unlikely. The mean estimate is simply to mitigate the polarity, but even so, no data reflects that a third of livestock here are on pasture, a third on solid, a third on liquid WMS. Improvements would take place in that portion of analysis specifically. In addition, the ICLEI protocol actually suggests using a percentage of animals whose manure is directed toward a WMS as a variable in calculations, and this is noted in the document. 

Thank you to Sara Isebrand at MSPC, Mauricio Leon, and Kristen Peterson for their contributions. 








